Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-08 aistats.tex
Pass over beginning of optim

% \hypersetup{draft}

@@

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, at the intersection of optimization and statistics, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta_*)},
\end{equation}
When $p(x\cond\theta)$ is an exponential family distribution,
$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.

@@

\paragraph{Maximum A Posteriori (MAP).}
Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat | \mD_n)
	%\propto p(\mD|\nat)p(\nat) 
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
\begin{align}
    \nabla \logpart(\MAPt) = \MAPm
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD_n$. 
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\subsection{MAP as Stochastic Mirror Descent}
The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
We recognize the update of Stochastic Mirror Descent -- a.k.a. Stochastic Bregman Gradient -- with step-size $\lr_n = \inv{n_0 + n}$.

@@

\subsection{Quadratic Case}
As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2} 
	=\frac{\Var(T)}{2n} 
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
\end{align}
so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

@@

{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
\begin{align}
    \bregmanconj(\m^* ; \m) 
    \leq \frac{L}{2} \norm{\m^* - \m}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.

@@

Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.

@@

\section{Optimization Perspective}
\fdk{Some sections are obv. too long, and some info irrelevant.}

In the context of optimization, why do we care?

\subsection{Stochastic mirror descent}

\fdk{
- Would mirror descent and the equivalence with MLE/MAP be introduced earlier?
- Mirror descent \citep{nemirovski1983problem,beck2003mirror}
- (also called NoLips \citep{bauschke2017descent}, 
relative gradient descent \citep{hanzely2018fastest},
Bregman proximal gradient, ...)
}

{\bf In online learning, } \fdk{bounded gradient assumption}

{\bf In smooth optimization, } there has been recent advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{bauschke2017descent,lu2018relatively}.
Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
relative to a reference function $A$ is defined 
as 
\aligns{
	\mu \cB_{A}(x, y)
	\leq 
	\cB_f(x,y)
	\leq 
	L \cB_A(x,y).
}
Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
Maximum likelihood in exponential families
that fits the theory developped for mirror descent well. 
Indeed, 
\aligns{
	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$,
and the the optimality gap is the divergence
\aligns{
	f(\theta) - f(\theta_*) = 
	\cB_A(\theta, \theta_*)
	= \KL(p(\cdot\cond\theta_*);\, p(\cdot\cond\theta)),
}
where $\theta_* = \arg\min_\theta f(\theta)$.
But the known results for stochastic mirror descent do not give convergence rates 
for all exponential families.

{\bf Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$ 
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex). 
This apply for the MLE problem of a Gaussian distribution 
with fixed variance, 
but not when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.

The work of \citet{bauschke2017descent,lu2018relatively}
has renewed interest for analyses of mirror descent in the smooth setting, 
and developping new tools to analyse problems
without relying on strong-convexity in an underlying norm.
However, they do not yet address the problem satisfactorily.

>>>>

\hypersetup{draft}

@@

But our understanding of stochastic optimization lacks in other domains. 
A generalization of the above problem, at the intersection of optimization and statistics, 
is the problem of minimizing 
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
\end{equation}
When $p(x\cond\theta)$ is an exponential family distribution,
$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.

@@

\paragraph{Maximum A Posteriori (MAP).}
Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat) 
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
\begin{align}
    \nabla \logpart(\MAPt) = \MAPm
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD_n$. 
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\subsection{MAP as Stochastic Mirror Descent}
\label{ssec:MAP=SMD}
The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
This parallel means it would be possible to obtain a finite convergence rate for the  MAP from an analysis of these algorithms. 
We give a special focus on this approach in Section~\ref{sec:optimization}.

@@

\subsection{Quadratic Case}
As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2} 
	=\frac{\Var(T)}{2n} 
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

@@

{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
\begin{align}
    \bregmanconj(\m^* ; \m) 
    \leq \frac{L}{2} \norm{\m^* - \m}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.

@@

\textbf{Remark:} In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.

@@

\section{Optimization Perspective}
\label{sec:optimization}
As we saw in Section~\ref{ssec:MAP=SMD}, MAP can be interpreted as Stochastic Mirror Descent (SMD). 
There are two consequences of this observation: 1. we may obtain a convergence rate for MAP from an optimization analysis, and 2. any insights gained from the MAP may inform further design and analysis of SMD.

\subsection{Relative Smoothness}
Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.

More recently, advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
relative to a reference function $A$ is defined  as 
\aligns{
	\mu \cB_{A}(x, y)
	\leq 
	\cB_f(x,y)
	\leq 
	L \cB_A(x,y).
}
Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
 Exponential families log-likelihood perfectly fits into this theory.
\aligns{
	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.
However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates 
for all exponential families.

{\bf Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$ 
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex). 
This apply for the MLE problem of a Gaussian distribution 
with fixed variance, 
but not when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.
Relative smoothness has renewed interest for analyses of mirror descent
without relying on strong-convexity in an underlying norm.
However, they do not yet address the problem satisfactorily.

\TODO{table of results of all 3 papers.}
